\huge{ \emph{Supplement to} Algorithms For Shaping a Particle Swarm\\ With a Shared Control Input Using Boundary Interaction}

### 2016-09-08 RA-L/theory.tex
Reducing size of pdf to less than 10 MB
compressed some images, made appendix able to be added easily.
some of Mary's Comments

\subsection{Using Boundaries: Fluid Settling In a Tank}\label{subsec:FluidInTank}
One method to control a swarm's shape in a bounded workspace is to simply push in a given direction until the swarm conforms to the boundary.
\paragraph{Square workplace}
%The formulas for means $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ are as follows: 
This section examines the mean $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ of a very large swarm of robots as they move inside a square workplace under the influence of gravity pointing in the direction $\beta$. The swarm is large, but the robots are small in comparison, and together occupy a constant area $A$. Under a global input such as gravity, they flow like water, moving to a side of the workplace and forming a polygonal shape, as shown in Fig.~\ref{fig:SquareFill}.

The range for the global input angle $\beta $ is [0,2$\pi $). In this range, the swarm assumes eight different polygonal shapes. The shapes alternate between triangles and trapezoids when the area $A$$<$1/2, and alternate between squares with one corner removed and trapezoids when $A$$>$1/2.

@@

The region of integration $R$ is the polygon containing the swarm. If the force angle is $\beta$, the mean when the swarm is in the lower-left corner is:
\begin{align}\label{eq:meanInSquareWorkspaceLL}
\bar{x}(A,\beta) &= \frac{\int_0^{\sqrt{2} \sqrt{-A \tan (\beta )}} \left(\int_0^{\sqrt{2} \sqrt{-A \cot (\beta )}+x \cot (\beta )} x \, dy\right) \, dx}{A} \nonumber \\
	&=\frac{1}{3} \sqrt{2} \sqrt{A \tan (\beta )}\\
\bar{y}(A,\beta) &= \frac{\int_0^{\sqrt{2} \sqrt{-A \tan (\beta )}} \left(\int_0^{\sqrt{2} \sqrt{-A \cot (\beta )}+x \cot (\beta )} y \, dy\right) \, dx}{A} \nonumber\\
	&=\frac{1}{3} \sqrt{2} \sqrt{A \cot (\beta )}
\end{align}
The full equations are included in the appendix, and are summarized in Fig.~\ref{fig:SquareFill}. A few highlights are that the correlation is maximized when the swarm is in a triangular shape, and is $\pm$1/2. The covariance of a triangle is always $\pm(A/18)$. Variance is minimized in the direction of $\beta$ and maximized orthogonal to $\beta$ when the swarm is in a rectangular shape. The range of mean positions are maximized when $A$ is small.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Circular workplace}
Though rectangular boundaries are common in artificial workspaces, biological workspaces are usually rounded.
Similar calculations can be computed for a circular workspace.  The workspace is a circle centered at (0,0) with radius 1 and thus area $\pi$.
For notational simplicity, the swarm is parameterized by the global control input signal $\beta$ and the fill-level $h$.  
Under a global input, the robot swarm fills the region under a chord with area
\begin{align}
A(h) = \arccos(1-h)-(1-h) \sqrt{(2-h) h}.
\end{align}
For a circular workspace, the locus of mean positions are aligned with $\beta$ and the mean position is at radius $r(h)$ from the center:
\begin{align}
r(h) = \frac{2 (-(h-2) h)^{3/2}}{3 \left(\sqrt{-(h-2) h} (h-1)+\arccos(1-h)\right)}
\end{align}
Variance $\sigma^2_x(\beta,h)$ is maximized at $\beta = \pi/2+n \pi$ and $h\approx1.43$, while covariance is maximized at $\beta = \pi3/4+n \pi$ and $h\approx0.92.$ For small $h$ values, correlation approaches $\pm1$. Results are summarized in Fig.~\ref{fig:CircleFill}.

@@

\subsection{Using Boundaries: Friction and Boundary Layers}\label{subsec:WallFriction}
Global inputs move a swarm uniformly.  
Controlling covariance requires breaking this uniform symmetry.  A swarm inside an axis-aligned rectangular workspace can reduce variance normal to a wall by simply pushing the swarm into the boundary. Directly controlling covariance by pushing the swarm into a boundary requires changes to the boundary.  An obstacle in the lower-right corner is enough to generate positive covariance.  Generating both positive and negative covariance requires additional obstacles.  Requiring special obstacle configuration also makes covariance control dependent on the local environment. 
  Instead of pushing our robots directly into a wall, this paper examines an oblique approach, by using boundaries that generate friction with the robots.  These frictional forces are  sufficient to break the symmetry caused by uniform inputs.  Robots touching a wall have a negative friction force that opposes movement along the boundary.  This  causes robots along the boundary to slow down compared to robots in free-space.

@@

\begin{align}
F_{\text{\emph{forward}}} &=  F \sin(\theta) - F_f  \nonumber \\
\text{where }  F_f &= \begin{cases}  \mu_f N, &  \mu_f N < F \sin(\theta)  \label{eq:frictionmodel}  \\
F \sin(\theta), & \text{else} \end{cases} \\ %\sign(F \sin(\theta) ) \cdot  \max(0, | F sin\theta |- |F_f|)
\text{and } N &= F \cos(\theta) \nonumber
\end{align}
 Fig.~\ref{fig:friction} shows the resultant forces on two robots when one is touching a wall. As illustrated, both experiences different net forces although each receives the same inputs.
  For ease of analysis, the following algorithms assume $\mu_f$ is infinite and robots touching the wall are prevented from sliding along the wall.
This means that if one robot is touching the wall and another robot is free, if the control input is parallel or into the wall, the touching robot will not move. 
There are many alternate models of friction that also break control symmetry. Fig.~\ref{fig:friction}c shows fluid flow along a boundary.  Fluid in the free-flow region moves uniformly, but flow decreases to zero in the boundary layer.  
\begin{align}
INSERT FORMULA \label{eq:boundarylayerflow}
\end{align}

@@

\section{Algorithms}\label{sec:algorithms}
\subsection{Position Control of $2$ Robots Using Wall Friction}\label{sec:PostionControl2Robots}
\begin{figure*}
\centering
\renewcommand{\figwid}{0.4\columnwidth}
{\begin{overpic}[width =\figwid]{S1_1.png}\put(10,15){$t$ = 0 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_2.png}\put(10,15){$t$ = 0.14 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_3.png}\put(10,15){$t$  = 0.29 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_4.png}\put(10,15){$t$  = 0.43 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_5.png}\put(10,15){$t$  = 0.76 s}
\end{overpic}}\\
%{\begin{overpic}[width =\figwid]{two_1.jpg}\put(45,75){$t$  = 0 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_2.jpg}\put(45,75){$t$  = 0.25 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_3.jpg}\put(45,75){$t$  = 0.5 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_4.jpg}\put(45,75){$t$  = 0.75 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_5.jpg}\put(45,75){$t$  = 1 s}
%\end{overpic}}\\
%\vspace{.5em}
%{\begin{overpic}[width =\figwid]{S3_1.jpg}\put(45,75){$t$ = 0 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_2.jpg}\put(45,75){$t$ = 0.25 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_3.jpg}\put(45,75){$t$  = 0.5 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_4.jpg}\put(45,75){$t$  = 0.75 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_5.jpg}\put(45,75){$t$  = 1 s}
%\end{overpic}}
%\vspace{-1em}
\caption{\label{fig:shapeControlMathematica1}{Frames from an implementation of Alg.\ \ref{alg:PosControl2Robots}: two robot positioning using walls with infinite friction. The algorithm only requires friction along the bottom and left walls.
Robot initial positions are shown by a crosshair, and final positions by a circled crosshair.  Dashed lines show the shortest route if robots could be controlled independently.  Solid arrows show path given by  Alg.\ \ref{alg:PosControl2Robots}.
Online demonstration and source code at \citep{Shahrokhi2015mathematicaParticle}.
%The bottom row shows an extreme case where the robots must switch position.
}
%\vspace{-2em}
}
\end{figure*}

This section describes Alg.~\ref{alg:PosControl2Robots}, which uses wall-friction to arbitrarily position two robots in a rectangular workspace.  This algorithm  introduces concepts that will be used for multi-robot positioning. It only requires collisions with two orthogonal walls, in this case, the bottom and left walls. Fig.~\ref{fig:shapeControlMathematica1} shows a Mathematica implementation of the algorithm, and is useful as a visual reference for the following description.

@@

\begin{figure}
\begin{center}
	\includegraphics[width=.9\columnwidth]{driftmove.pdf}
\end{center}
\vspace{-1em}
\caption{\label{fig:driftmove}
A  $\operatorname{DriftMove}(\alpha, \beta, \epsilon)$ to the right repeats a triangular movement sequence $\{ (\beta/2,-\epsilon),(\beta/2,\epsilon),(-\alpha,0)\}$. Robot $A$ touching a top wall moves right $\beta$ units, while robots not touching the top move right $\beta-\alpha$.
}
\end{figure}
\begin{figure}
\begin{center}
	\includegraphics[width=1.0\columnwidth]{multirobotSliderHardware.pdf}
\end{center}
\vspace{-1em}
\caption{\label{fig:construction2d}
Illustration of Alg.\ \ref{alg:PosControlNRobots}, $n$ robot position control  using wall friction.
}
\end{figure}

>>>>

\subsection{Using Boundaries: Fluid Settling In a Tank}\label{subsec:FluidInTank}
One method to control a swarm's shape in a bounded workspace is to simply push in a given direction until the swarm conforms to the boundary.
\paragraph{Square workplace}
%The formulas for means $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ are as follows: 
We first examine the mean $(\bar{x},\bar{y})$, covariance $(\sigma^2_x,\sigma^2_y,\sigma_{xy})$, and correlation $\rho_{xy}$ of a very large swarm of robots as they move inside a square workplace under the influence of gravity pointing in the direction $\beta$. The swarm is large, but the robots are small in comparison, and together occupy a constant area $A$. Under a global input such as gravity, they flow like water, moving to a side of the workplace and forming a polygonal shape, as shown in Fig.~\ref{fig:SquareFill}.

The range for the global input angle $\beta $ is [0,2$\pi $). In this range, the swarm assumes eight different polygonal shapes. Therse shapes alternate between triangles and trapezoids when the area $A$$<$1/2, and between squares with one corner removed and trapezoids when $A$$>$1/2.

@@

The region of integration $R$ is the polygon containing the swarm. If the force angle is $\beta$, the mean when the swarm is in the lower-left corner is:
\begin{align}\label{eq:meanInSquareWorkspaceLL}
\bar{x}(A,\beta) &= \frac{\int_0^{\sqrt{2} \sqrt{-A \tan (\beta )}} \left(\int_0^{\sqrt{2} \sqrt{-A \cot (\beta )}+x \cot (\beta )} x \, dy\right) \, dx}{A} \nonumber \\
	&=\frac{ \sqrt{2}}{3} \sqrt{A \tan (\beta )}\\
\bar{y}(A,\beta) &= \frac{\int_0^{\sqrt{2} \sqrt{-A \tan (\beta )}} \left(\int_0^{\sqrt{2} \sqrt{-A \cot (\beta )}+x \cot (\beta )} y \, dy\right) \, dx}{A} \nonumber\\
	&=\frac{\sqrt{2}}{3}  \sqrt{A \cot (\beta )}
\end{align}
The full equations are included in the appendix, and are summarized in Fig.~\ref{fig:SquareFill}. A few highlights are that the correlation is maximized $\pm$1/2 when the swarm is in a triangular shape. The covariance of a triangle is always $\pm(A/18)$. Variance is minimized in the direction of $\beta$ and maximized orthogonal to $\beta$ when the swarm is in a rectangular shape. The range of mean positions are maximized when $A$ is small.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Circular workplace}
Though rectangular boundaries are common in artificial workspaces, biological workspaces are usually rounded.
Similar calculations can be made for a circular workspace.  The workspace is a circle centered at (0,0) with radius 1 and thus area $\pi$.
For notational simplicity, the swarm is parameterized by the global control input signal $\beta$ and the fill-level $h$.  
Under a global input, the robot swarm fills the region under a chord with area
\begin{align}
A(h) = \arccos(1-h)-(1-h) \sqrt{(2-h) h}.
\end{align}
For a circular workspace, the locus of mean positions are aligned with $\beta$ and the mean position is at radius $r(h)$ from the center:
\begin{align}
r(h) = \frac{2 (-(h-2) h)^{3/2}}{3 \left(\sqrt{-(h-2) h} (h-1)+\arccos(1-h)\right)}
\end{align}
Variance $\sigma^2_x(\beta,h)$ is maximized at $\beta = \pi/2+n \pi$ and $h\approx1.43$, while covariance is maximized at $\beta = \pi3/4+n \pi$ and $h\approx0.92.$ For small $h$ values, correlation approaches $\pm1$. Results are displayed in Fig.~\ref{fig:CircleFill}.

@@

\subsection{Using Boundaries: Friction and Boundary Layers}\label{subsec:WallFriction}
Global inputs move a swarm uniformly.  
Controlling covariance requires breaking this uniform symmetry.  A swarm inside an axis-aligned rectangular workspace can reduce variance normal to a wall by simply pushing the swarm into the boundary. Directly controlling covariance by pushing the swarm into a boundary requires changes to the boundary.  An obstacle in the lower-right corner is enough to generate positive covariance, but generating both positive and negative covariance requires additional obstacles.  Requiring a special obstacle configuration also makes covariance control dependent on the local environment. 
  Instead of pushing our robots directly into a wall, this paper examines an oblique approach, by using boundaries that generate friction with the robots.  These frictional forces are  sufficient to break the symmetry caused by uniform inputs.  Robots touching a wall have a negative friction force that opposes movement along the boundary.  This  causes robots along the boundary to move more slowly than robots in free-space.

@@

\begin{align}
F_{\text{\emph{forward}}} &=  F \sin(\theta) - F_f  \nonumber \\
\text{where }  F_f &= \begin{cases}  \mu_f N, &  \mu_f N < F \sin(\theta)  \label{eq:frictionmodel}  \\
F \sin(\theta), & \text{else} \end{cases} \\ %\sign(F \sin(\theta) ) \cdot  \max(0, | F sin\theta |- |F_f|)
\text{and } N &= F \cos(\theta) \nonumber
\end{align}
 Fig.~\ref{fig:friction} shows the resultant forces on two robots when one is touching a wall. As illustrated, both experience different net forces although each receives the same inputs.
  For ease of analysis, the following algorithms assume $\mu_f$ is infinite and robots touching the wall are prevented from sliding along the wall.
This means that if one robot is touching the wall and another robot is free, the touching robot will not move when the control input is parallel or into the wall. 
There are many alternate models of friction that also break control symmetry. Fig.~\ref{fig:friction}c shows fluid flow along a boundary.  Fluid in the free-flow region moves uniformly, but flow decreases to zero in the boundary layer.  
\begin{align}
u(y) = u_0 [1- \frac{(y-h)^2}{h^2} = u_0 \frac{y}{h} [2- \frac{y}{h}] \label{eq:boundarylayerflow}
\end{align}

@@

\section{Algorithms}\label{sec:algorithms}
\subsection{Position Control of $2$ Robots Using Wall Friction}\label{sec:PostionControl2Robots}
\begin{figure*}
\centering
\renewcommand{\figwid}{0.4\columnwidth}
{\begin{overpic}[width =\figwid]{S1_1.png}\put(10,15){$t$ = 0 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_2.png}\put(10,15){$t$ = 0.14 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_3.png}\put(10,15){$t$  = 0.29 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_4.png}\put(10,15){$t$  = 0.43 s}
\end{overpic}
\begin{overpic}[width =\figwid]{S1_5.png}\put(10,15){$t$  = 0.76 s}
\end{overpic}}\\
%{\begin{overpic}[width =\figwid]{two_1.jpg}\put(45,75){$t$  = 0 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_2.jpg}\put(45,75){$t$  = 0.25 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_3.jpg}\put(45,75){$t$  = 0.5 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_4.jpg}\put(45,75){$t$  = 0.75 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{two_5.jpg}\put(45,75){$t$  = 1 s}
%\end{overpic}}\\
%\vspace{.5em}
%{\begin{overpic}[width =\figwid]{S3_1.jpg}\put(45,75){$t$ = 0 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_2.jpg}\put(45,75){$t$ = 0.25 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_3.jpg}\put(45,75){$t$  = 0.5 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_4.jpg}\put(45,75){$t$  = 0.75 s}
%\end{overpic}
%\begin{overpic}[width =\figwid]{S3_5.jpg}\put(45,75){$t$  = 1 s}
%\end{overpic}}
%\vspace{-1em}
\caption{\label{fig:shapeControlMathematica1}{Frames from an implementation of Alg.\ \ref{alg:PosControl2Robots}: two robot positioning using walls with infinite friction. The algorithm only requires friction along the bottom and right walls.
Robot initial positions are shown by a crosshair, and final positions by a circled crosshair.  Dashed lines show the shortest route if robots could be controlled independently.  Solid arrows show path given by  Alg.\ \ref{alg:PosControl2Robots}.
Online demonstration and source code at \citep{Shahrokhi2015mathematicaParticle}.
%The bottom row shows an extreme case where the robots must switch position.
}
%\vspace{-2em}
}
\end{figure*}

Alg.~\ref{alg:PosControl2Robots} uses wall-friction to arbitrarily position two robots in a rectangular workspace.  This algorithm  introduces concepts that will be used for multi-robot positioning. It only requires collisions with two orthogonal walls, in this case, the bottom and left walls. Fig.~\ref{fig:shapeControlMathematica1} shows a Mathematica implementation of the algorithm, and is useful as a visual reference for the following description.

@@

\begin{figure}
\begin{center}
	\includegraphics[width=.9\columnwidth]{driftmove.pdf}
\end{center}
\vspace{-1em}
\caption{\label{fig:driftmove}
A  $\operatorname{DriftMove}(\alpha, \beta, \epsilon)$ to the right repeats a triangular movement sequence $\{ (\beta/2,-\epsilon),(\beta/2,\epsilon),(-\alpha,0)\}$. Robot $A$ touching a top wall moves right $\beta$ units, while robots not touching the top move right $\beta-\alpha$.
}
\end{figure}
\begin{figure}
\begin{center}
	\includegraphics[width=1.0\columnwidth]{multirobotSliderHardwareComp.pdf}
\end{center}
\vspace{-1em}
\caption{\label{fig:construction2d}
Illustration of Alg.\ \ref{alg:PosControlNRobots}, $n$ robot position control  using wall friction.
}
\end{figure}
